Non-Vertical Cultural Transmission, Assortment, and the Evolution of Cooperation

### 2021-07-23 ms/Thesis/thesis.tex
formatting; ackwoledgements

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{lipsum}  
\usepackage[title,page]{appendix}
\usepackage{chngcntr}
%\usepackage{endfloat}

@@

% Line Spacing 
\linespread{1.25}

% Species names
%% Meta-Command for defining new species macros
\usepackage{xspace}

\newcommand{\species}[3]{%
  \newcommand{#1}{\gdef#1{\textit{#3}\xspace}\textit{#2}\xspace}}

\species{\yeast}{Saccharomyces cerevisiae}{S.~cerevisiae}
\species{\calbicans}{Candida albicans}{C.~albicans}
\species{\cneoformans}{Cryptococcus neoformans}{C.~neoformans}

% line numbers
\usepackage[displaymath, mathlines]{lineno}
\renewcommand\linenumberfont{\normalfont\small\sffamily}
\linenumbers
\modulolinenumbers[2]

@@

% Title page
\title{Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation}

@@

	\vspace{1.5cm}
	{\Huge Non-Vertical Cultural 
  Transmission, 
  Assortment, and the 
  Evolution of 
  Cooperation \par}
	\vspace{1.5cm}
	{\large by\par}
	{\Large\bfseries Dor Cohen\par}
	\vspace{1cm}
	{\large Supervised by Dr. Yoav Ram\par}
	{School of Zoology, Faculty of Life Sciences, Tel Aviv University\par}

@@

	\vspace{1cm}
	% Bottom of the page
	{\large \monthyeardate\today \par}

@@

{\huge \begin{center} \textbf{Abstract} \end{center}}

Cultural evolution of cooperation under vertical and non-vertical cultural transmission is studied, and conditions are found for fixation and coexistence of cooperation and defection. 
The evolution of cooperation is facilitated by its horizontal transmission and by an association between social interactions and horizontal transmission.
The effect of oblique transmission depends on the horizontal transmission bias.
Stable polymorphism of cooperation and defection can occur, and
when it does, reduced association between social interactions and horizontal transmission evolves, which leads to a decreased frequency of cooperation and lower population mean fitness.
The deterministic conditions are compared to outcomes of stochastic simulations of structured populations.
Parallels are drawn with Hamilton's rule incorporating relatedness and assortment.

\pagebreak

@@

There is an ongoing debate about to what extent kin selection explains evolution of cooperation and altruism.
It has been suggested that kin selection to explain the cooperative behaviour of eusocial insects like the honey bee.
The most significant argument against kin selection is that cooperation can evolve with zero relatedness \cite{wilson2005kin}. This makes Hamilton's rule incomplete according to Wilson \cite{wilson2005kin}. Foster et al. \cite{foster2006kin} reject this claim. 
They argue that altruism without relatedness can not evolve. They refer us to Hamilton who claimed that relatedness can arise without recent common ancestry. 
Wilson also criticises kin selection on the grounds that environmental or ecological factors probably be more important than relatedness in determining social actions. On the other hand, Foster et al. \cite{foster2006kin} argue that kin selection does not ignore ecology. 
Hamilton’s rule shows that environmental factors causing a high benefit: cost ratio will favour cooperation.

Beside kin selection, two other major theroies were suggested to explain to evolution of cooperation.

@@

\textbf{Group Selection} theory posits that cooperation is favoured because of the advantage to the whole group, if selection acts at the group level in addition to the individual level. A common model for group selection work as is: the population is divided into groups. In each group there are cooperators, which help other group members, and defectors, which do not help. 
Individuals reproduce proportional to their fitness. Offspring are added to the same group.
If a group reaches a certain size it can split to two groups. A group that grows faster will split more often. Groups of cooperators grow faster than groups of defectors.
Therefore, cooperation can evolve in this model when
\begin{equation} \label{groupselection}
\frac{b}{c}>1+\frac{n}{m}\, ,
\end{equation}
where \emph{n} is the maximum group size and \emph{m} is the number of groups.
\\

@@

\citet{Eshel1982} studied a related model for the evolution of cooperative behavior.
Their model included \emph{assortative meeting}, or non-random encounters, where a fraction $m$ of individuals in the population each interact specifically with an individual of the same phenotype, and a fraction $1-m$ interacts  with a randomly chosen individual.  
Such assortative meeting may be due, for example, to population structure or active partner choice.
In their model, cooperative behavior can evolve if
\citep[eq.~3.2]{Eshel1982}
\begin{equation} \label{eq:eshel1982}
c < b \cdot m \,,
\end{equation}
where $b$ and $c$ are the benefit and cost of cooperation\footnote{In an extended model, which allows an individual to encounter $N$ individuals before choosing a partner, the right hand side is multiplied by $E[N]$, the expected number of encounters \citep[eq.~4.6]{Eshel1982}.
}.

@@

To understand the evolution of cooperation we are going to use \emph{replicator dynamics}. The replicator in replicator dynamics has the ability to make one or more copies of itself.
The replicator can be a gene, a phenotype, a strategy in a game and etc. In evolutionary game theory context replicator is a different strategy in the game. For example, in the case of cooperation whether the individual is a cooperative or a defector.
In replicator dynamics we assume large population of replicators, which interact with respect to their frequency.
Interactions of different replicator affect the fitness acorrding to some payoff matrix. This payoff matrix depands on the game which is played. 
The most common game to describe cooperation is the \emph{Prisoner's Dilemma}\citep{nowak1992tit}.
Similar to dominant strategies bringing forth Nash equilibria when games are repeated, strategies in replicator dynamics can become evolutionary stable. Such strategies are called \emph{evolutionarily stable strategies (ESS)}.
Such strategies cannot be invaded by any other strategy that is initially rare. Evolutionarily stable strategies maximize the expected fitness of its replicators and therefore, maximize the population mean fitness.
A fundamental question is to test if natural selection, operating within the framework of known genetical models, leads to ESS.
The answer to the question above is that natural selection in many cases does not lead to an increase in the population mean fitness. Therefore, it is not generally true that natural selection does operate as to maximize the average individual fitness resulting in an ESS.
\citet{sacks1967stable} even showed that natural selection at a diploid locus model, when relative fitnesses of the different genotypes are functions of gene frequency, leads to stable equilibrium gene frequencies that corresponds to not maximum mean fitness.
Instead of searching for evolutionarily stable strategies, we focus on \emph{evolutionary genetic stblility (EGS)}\citep{lessard1990evolutionary}. EGS occurs when the frequencies of types (genotypes, phenotypes etc.) remains stable.
Here, we study both \emph{local stability} and \emph{external stability}. \emph{Local stability} occurs when a system near the equilibrium wiil approach it.
\emph{External stability} \citep{Liberman1986modifiers,Altenberg2017} is local stability of the equilibrium to invasion by a modifier allele \emph{m}. One of the main questions in the evolution of cooperation is under what conditions such invasions are possible, or in other words under what conditions the system is externally unstable?

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Models
\newpage
\section{Models}

\subsection{Basic Model}
\label{section:basicmodel}
Consider a large population whose members can have one of two phenotypes: $\phi=A$ for cooperators or $\phi=B$ for defectors.
An offspring inherits its phenotype from its parent via vertical transmission with probability $v$ or from a random individual in the parental population via oblique transmission with probability $(1-v)$ (\autoref{fig:illustration}a). 
Following~\citet{ram2018evolution}, given that the parent's phenotype is $\phi$ and assuming uni-parental inheritance~\citep{Zefferman2016}, the conditional probability that the phenotype $\phi'$ of the offspring is $A$ is 
\begin{equation} \label{eq:vertical_oblique_transmission}
P(\phi'=A \mid \phi) = \begin{cases}
v + (1-v)p, & \text{if } \phi=A \\
(1-v)p, & \text{if } \phi=B
\end{cases},
\end{equation}
where $p=P(\phi=A)$ is the frequency of $A$ among all adults in the parental generation.

@@

\subsection{Invasion Model}
In this section, we extend the model described in \autoref{section:basicmodel}. 
We assume coexistence of both phenotypes (cooperators and defectors) and that the frequency of each phenotye stays stable over time.
Hence, we assume that the we start from a stable equilibrium. Denote the frequency of cooperators in the population as $\hat{p}^*$ and frequency of defectors as $1-\hat{p}^*$.

@@

%%% Figure: model illustration - invasion model
\begin{figure}[pt]
  \centering
  \includegraphics[width=0.7\textwidth]{../PRSB_figures/fig_2.pdf}
  \caption{\textbf{Invasion model illustration.} Starting from stable equilibrium with coexistence of both phenotypes. 
  A mutation that affected only the interaction-transmission association $\alpha$ invades the population }
  \label{fig:invasion_illustration}
\end{figure}

\newpage
\subsection{Population structure model}
\label{sec:spatialModel}
Interaction-transmission association may also emerge from population structure.
Consider a  population colonizing a two-dimensional grid of size 100-by-100, where each site is inhabited by one individual, similarly to the model of \citet{lewin2020rockpaperscissors}.
Each individual is characterized by its phenotype: either cooperator, $A$, or defector, $B$.
Initially, each site in the grid is randomly colonized by either a cooperator or a defector, with equal probability.
In each generation, half of the individuals are randomly chosen to "initiate" interactions, and these
initiators interact with a random neighbor (i.e. individual in a neighboring site) in a Prisoner's Dilemma game (\autoref{fig:illustration}c) and a random neighbor (with replacement) for horizontal cultural transmission (\autoref{fig:illustration}b).
The expected number of each of these interactions per individual per generation is one, but the realized number of interactions can be zero, one, or even more than one, and in every interaction both individuals are affected, not just the initiator.
The effective interaction-transmission association $\alpha$ in this model is the probability that the same neighbor is picked for both interactions, or $\alpha=1/M$, where $M$ is the number of neighbors.
On an infinite grid, $M=8$ (i.e. Moore neighbourhood~\citep{moore1962machine}), but on a finite grid $M$ can be lower in neighbourhoods close to the grid border.
As before, $T_A$ and $T_B$ are the probabilities of successful horizontal transmission of phenotypes $A$ and $B$, respectively.

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Results
\newpage
\section{Results}

@@

It is interesting to examine the general effect of interaction-transmission association $\alpha$ on the evolution of cooperation.
Define the interaction-transmission association thresholds, $a_1$ and $a_2$, as 
\begin{equation} \label{eq:boundries_assortative_meeting_general_case}
  a_1 = \frac{c\cdot v(1-T_A) -(T_A-T_B)(1+b-c)}{b\cdot v \cdot T_B}, \quad
  a_2 = \frac{c\cdot v(1-T_B)-(T_A-T_B)}{b\cdot v\cdot T_A} \;.
\end{equation}

@@

\begin{figure}[p]
  \centering       
    \includegraphics[width=1\textwidth]{../PRSB_figures/fig_3.pdf}
    \caption{\textbf{Evolution of cooperation under vertical, oblique, and horizontal cultural transmission.} 
    The figure shows parameter ranges for global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable polymorphism (blue), and stable polymorphism of cooperation and defection (yellow).
	In all cases the vertical transmission rate $v$ is on the x-axis.
	(\textbf{a-b})~Cost of cooperation $c$ is on the y-axis and the cost thresholds $\gamma_1$ and $\gamma_2$ (Eqs.~\ref{eq:cost_thresholds_v_threshold}) are represented by the solid and dashed lines, respectively. 
    (\textbf{c-d})~Interaction-transmission association $\alpha$ is on the y-axis and the interaction-transmission association thresholds $a_1$ and $a_2$ (Eqs.~\ref{eq:boundries_assortative_meeting_general_case}) are represented by the solid and dashed lines, respectively. 
    Horizontal transmission is biased in favor of cooperation, $T_A>T_B$, in (\textbf{a}) and (\textbf{c}), or defection, $T_A<T_B$, in (\textbf{b}) and (\textbf{d}).    
    Here, $T_A = 0.5$, and
    (\textbf{a}) $b=1.2$, , $T_B = 0.4$, $\alpha = 0.4$;
    (\textbf{b}) $b=2$, $T_B = 0.7$, $\alpha = 0.7$;
    (\textbf{c}) $b=1.2$, $T_B = 0.4$, $c=0.5$;
    (\textbf{d}) $b=2$, $T_B = 0.7$, $c=0.5$.
    }
    \label{fig:equilibria}
\end{figure}

@@

\begin{figure}[p]
  \centering       
	\includegraphics[width=1\textwidth]{../PRSB_figures/fig_4.pdf}
    \caption{\textbf{Evolution of cooperation under vertical and horizontal cultural transmission (v=1).} 
    The figure shows parameter ranges for global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable polymorphism (blue), and stable polymorphism of cooperation and defection (yellow).
    (\textbf{a-c})~The horizontal transmission bias ($T_A-T_B$) is on the x-axis.
    In panels \textbf{(a)} and \textbf{(b)}, the cost of cooperation $c$ is on the y-axis and the cost thresholds $\gamma_1$ and $\gamma_2$ (\autoref{eq:cost_thresholds_v_threshold}) are the solid and dashed lines, respectively. 
	In panel \textbf{(c)}, interaction-transmission association $\alpha$ is on the y-axis and the interaction-transmission association thresholds $a_1$ and $a_2$ (Eqs.~\ref{eq:boundries_assortative_meeting_general_case}) are the solid and dashed lines, respectively.
    Here, $b=1.3$, $T_A=0.4$, $v=1$, (a)~$\alpha = 0$, (b)~$\alpha = 0.7$, (c)~$c = 0.35$.    
    \textbf{(d)}~Change in frequency of cooperation among juveniles ($\hat{p}'-\hat{p}$) as a function of the frequency ($\hat{p}$), see \autoref{eq:nextgen_juveniles}.
    The orange curve shows convergence to a stable polymorphism ($T_A = 0.4$, $T_B = 0.9$, $b = 12$, $c=0.35$, $v=1$, and $\alpha = 0.45$). % todo ref eq 12
    The blue curve shows fixation of either cooperation or defection, depending on the initial frequency ($T_A = 0.5$, $T_B = 0.1$, $b = 1.3$, $c=0.904$, $v=1$, and $\alpha = 0.4$).
    Black circles show the three equilibria.
  	}
    \label{fig:equilibria_v1}
\end{figure}
\newpage

@@

\begin{figure}[h]
  \centering       
    \includegraphics[scale=0.5]{../PRSB_figures/fig_5.pdf}
    \caption{\textbf{Unstable and stable equilibrium dynamics} 
    The figure shows the frequency of cooperation over time (generations) starting from different initial frequencies.
    In \textbf{(a)} the system has unstable equilibrium and therefore, all the curves are moving away from it.
    In \textbf{(b)} the system has stable equilibrium and therefore, all the curves moves towards it. 
    Here, $T_A = 0.5$, $c = 0.5$,  $v=0.59$ and \textbf{(a)} $T_B = 0.4$, $\alpha=0.1$ and $b=1.2$. \textbf{(b)} $T_B=0.7$, $\alpha=0.6$ and $b=2.3$.}
    \label{fig:time_series}
\end{figure}

@@

%%% Figure: Invasion
\begin{figure}[h]
  \centering
  \includegraphics[width=1\textwidth]{../PRSB_figures/fig_6.pdf}
  \caption{
  \textbf{Reduction principle for interaction-transmission association.} 
  Consecutive fixation of modifier alleles that reduce interaction-transmission association $\alpha$ in numerical simulations of evolution with two modifier alleles (Eq.~D1).
  When an invading modifier allele is established in the population (frequency > 99.95\%), a new modifier allele that reduces interaction-transmission association by 5\% is introduced (at initial frequency 0.5\%).
  \textbf{(a)}~The frequency of the cooperative phenotype $A$ over time.
  \textbf{(b)}~The frequency of the invading modifier allele $m$ over time. 
  \textbf{(c)}~The population mean fitness ($\bar{w}$) over time.
  Here, $v=1$, $c = 0.05$, $b=1.3$, $T_A=0.4<T_B=0.7$, initial interaction-transmission association $\alpha_1=0.7$, lower interaction-transmission association threshold $a_2=0.605$.  
  }
  \label{fig:invasion}
\end{figure}
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Population structure}
Here, we are going to run stochastic simulations to learn more about evolution of Cooperation when the population is structured. 
The model is described in \autoref{sec:spatialModel}.
All the simulations in this section were made by Ohad Lewin-Epstein from Tel Aviv University.
The outcomes of stochastic simulations with such a structured population are shown in \autoref{fig:spatial}, which demonstrates that the highest cost of cooperation $c$ that permits the evolution of cooperation agrees with the conditions derived above for our model without population structure or stochasticity.
An example of stochastic stable polymorphism is shown in \autoref{fig:spatial}c.
Changing the simulation so that selection is local (i.e. sites can only be settled by offspring of neighboring parents) had only a minor effect on the agreement with the derived conditions (\autoref{fig:spatial_local_selection}).

@@

\newpage

>>>>

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{lipsum}  
\usepackage[title,page,titletoc]{appendix}
\usepackage{chngcntr}
%\usepackage{endfloat}
\usepackage[nottoc,numbib]{tocbibind} % bibliography in toc

@@

% Line Spacing 
\linespread{1.5}

% line numbers
%\usepackage[displaymath, mathlines]{lineno}
%\renewcommand\linenumberfont{\normalfont\small\sffamily}
%\linenumbers
%\modulolinenumbers[2]

@@

% Title page
\title{Non-Vertical Cultural Transmission, Assortment, and the Evolution of Cooperation}

@@

	\vspace{1.5cm}
	{\Huge Non-Vertical Cultural 
  Transmission, 
  Assortment, and the 
  Evolution of 
  Cooperation \par}
%	\vspace{1cm}
	{\large by\par}
	{\Large\bfseries Dor Cohen\par}
	\vspace{1cm}
	{\large Supervised by Dr. Yoav Ram\par}
	{School of Zoology, Faculty of Life Sciences, Tel Aviv University\par}

@@

	\vspace{1cm}
	% Bottom of the page
	{\large \today \par}

@@

% Abstract
\begin{abstract}
Cultural evolution of cooperation under vertical and non-vertical cultural transmission is studied, and conditions are found for fixation and coexistence of cooperation and defection. 
The evolution of cooperation is facilitated by its horizontal transmission and by an association between social interactions and horizontal transmission.
The effect of oblique transmission depends on the horizontal transmission bias.
Stable polymorphism of cooperation and defection can occur, and
when it does, reduced association between social interactions and horizontal transmission evolves, which leads to a decreased frequency of cooperation and lower population mean fitness.
The deterministic conditions are compared to outcomes of stochastic simulations of structured populations.
Parallels are drawn with Hamilton's rule incorporating relatedness and assortment.
\end{abstract}

%\pagebreak

@@

There is an ongoing debate about to what extent kin selection explains evolution of cooperation and altruism.
It has been suggested that kin selection to explain the cooperative behaviour of eusocial insects like the honey bee.
The most significant argument against kin selection is that cooperation can evolve with zero relatedness \cite{wilson2005kin}. This makes Hamilton's rule incomplete according to Wilson \cite{wilson2005kin}. Foster et al. \cite{foster2006kin} reject this claim. 
They argue that altruism without relatedness can not evolve. They refer us to Hamilton who claimed that relatedness can arise without recent common ancestry. 
Wilson also criticises kin selection on the grounds that environmental or ecological factors probably be more important than relatedness in determining social actions. On the other hand, Foster et al. \cite{foster2006kin} argue that kin selection does not ignore ecology. 
Hamilton's rule shows that environmental factors causing a high benefit: cost ratio will favour cooperation.\\
Beside kin selection, two other major theroies were suggested to explain to evolution of cooperation.

@@

\textbf{Group Selection} theory posits that cooperation is favoured because of the advantage to the whole group, if selection acts at the group level in addition to the individual level. A common model for group selection work as is: the population is divided into groups. In each group there are cooperators, which help other group members, and defectors, which do not help. 
Individuals reproduce proportional to their fitness. Offspring are added to the same group.
If a group reaches a certain size it can split to two groups. A group that grows faster will split more often. Groups of cooperators grow faster than groups of defectors.
Therefore, cooperation can evolve in this model when
\begin{equation} \label{groupselection}
\frac{b}{c}>1+\frac{n}{m}\, ,
\end{equation}
where \emph{n} is the maximum group size and \emph{m} is the number of groups.

@@

\citet{Eshel1982} studied a related model for the evolution of cooperative behavior.
Their model included \emph{assortative meeting}, or non-random encounters, where a fraction $m$ of individuals in the population each interact specifically with an individual of the same phenotype, and a fraction $1-m$ interacts  with a randomly chosen individual.  
Such assortative meeting may be due, for example, to population structure or active partner choice.
In their model, cooperative behavior can evolve if~\citep[eq.~3.2]{Eshel1982}
\begin{equation} \label{eq:eshel1982}
c < b \cdot m \,,
\end{equation}
where $b$ and $c$ are the benefit and cost of cooperation\footnote{In an extended model, which allows an individual to encounter $N$ individuals before choosing a partner, the right hand side is multiplied by $E[N]$, the expected number of encounters \citep[eq.~4.6]{Eshel1982}.
}.

@@

To understand the evolution of cooperation we are going to use \emph{replicator dynamics}. The replicator in replicator dynamics has the ability to make one or more copies of itself.
The replicator can be a gene, a phenotype, a strategy in a game and etc. In evolutionary game theory context replicator is a different strategy in the game. For example, in the case of cooperation whether the individual is a cooperative or a defector.
In replicator dynamics we assume large population of replicators, which interact with respect to their frequency.
Interactions of different replicator affect the fitness acorrding to some payoff matrix. This payoff matrix depands on the game which is played. 
The most common game to describe cooperation is the \emph{Prisoner's Dilemma}\citep{nowak1992tit}.
Similar to dominant strategies bringing forth Nash equilibria when games are repeated, strategies in replicator dynamics can become evolutionary stable. Such strategies are called \emph{evolutionarily stable strategies (ESS)}.
Such strategies cannot be invaded by any other strategy that is initially rare. Evolutionarily stable strategies maximize the expected fitness of its replicators and therefore, maximize the population mean fitness.
A fundamental question is to test if natural selection, operating within the framework of known genetical models, leads to ESS.
The answer to the question above is that natural selection in many cases does not lead to an increase in the population mean fitness. Therefore, it is not generally true that natural selection does operate as to maximize the average individual fitness resulting in an ESS.
\citet{sacks1967stable} even showed that natural selection at a diploid locus model, when relative fitnesses of the different genotypes are functions of gene frequency, leads to stable equilibrium gene frequencies that corresponds to not maximum mean fitness.
Instead of searching for evolutionarily stable strategies, we focus on \emph{evolutionary genetic stblility (EGS)}~\citep{lessard1990evolutionary}.
EGS occurs when the frequencies of types (genotypes, phenotypes etc.) remains stable.
Here, we study both \emph{local stability} and \emph{external stability}. \emph{Local stability} occurs when a system near the equilibrium wiil approach it.
\emph{External stability} \citep{Liberman1986modifiers,Altenberg2017} is local stability of the equilibrium to invasion by a modifier allele \emph{m}. One of the main questions in the evolution of cooperation is under what conditions such invasions are possible, or in other words under what conditions the system is externally unstable?

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Models
%\newpage
\section{Models}

\subsection{Basic model}
\label{section:basicmodel}
Consider a large population whose members can have one of two phenotypes: $\phi=A$ for cooperators or $\phi=B$ for defectors.
An offspring inherits its phenotype from its parent via vertical transmission with probability $v$ or from a random individual in the parental population via oblique transmission with probability $(1-v)$ (\autoref{fig:illustration}a). 
Following~\citet{ram2018evolution}, given that the parent's phenotype is $\phi$ and assuming uni-parental inheritance~\citep{Zefferman2016}, the conditional probability that the phenotype $\phi'$ of the offspring is $A$ is 
\begin{equation} \label{eq:vertical_oblique_transmission}
P(\phi'=A \mid \phi) = \begin{cases}
v + (1-v)p, & \text{if } \phi=A \\
(1-v)p, & \text{if } \phi=B
\end{cases},
\end{equation}
where $p=P(\phi=A)$ is the frequency of $A$ among all adults in the parental generation.

@@

\subsection{Modifier model}
In this section, we extend the model described in \autoref{section:basicmodel}. 
We assume coexistence of both phenotypes (cooperators and defectors) and that the frequency of each phenotype stays stable over time.
Hence, we assume that the we start from a stable equilibrium. Denote the frequency of cooperators in the population as $\hat{p}^*$ and frequency of defectors as $1-\hat{p}^*$.

@@

%%% Figure: model illustration - invasion model
\begin{figure}[pt]
  \centering
  \includegraphics[width=0.7\textwidth]{../PRSB_figures/fig_2.pdf}
  \caption{\textbf{Modifier model illustration.} Starting from stable equilibrium with coexistence of both phenotypes (red and green) and the resident modifier allele with interaction-transmission association $\alpha_1$.
  An invading modifier allele that modifies the interaction-transmission association to $\alpha_2$ invades the population.
  }
  \label{fig:invasion_illustration}
\end{figure}

\newpage
\subsection{Spatial model}
\label{sec:spatialModel}
Interaction-transmission association may also emerge from population structure.
Consider a  population colonizing a two-dimensional grid of size 100-by-100, where each site is inhabited by one individual, similarly to the model of \citet{lewin2020rockpaperscissors}.
Each individual is characterized by its phenotype: either cooperator, $A$, or defector, $B$.
Initially, each site in the grid is randomly colonized by either a cooperator or a defector, with equal probability.
In each generation, half of the individuals are randomly chosen to "initiate" interactions, and these
initiators interact with a random neighbor (i.e. individual in a neighboring site) in a Prisoner's Dilemma game (\autoref{fig:illustration}c) and a random neighbor (with replacement) for horizontal cultural transmission (\autoref{fig:illustration}b).
The expected number of each of these interactions per individual per generation is one, but the realized number of interactions can be zero, one, or even more than one, and in every interaction both individuals are affected, not just the initiator.
The effective interaction-transmission association $\alpha$ in this model is the probability that the same neighbor is picked for both interactions, or $\alpha=1/M$, where $M$ is the number of neighbors.
On an infinite grid, $M=8$ (i.e. Moore neighbourhood~\citep{moore1962machine}), but on a finite grid $M$ can be lower in neighbourhoods close to the grid border.
As before, $T_A$ and $T_B$ are the probabilities of successful horizontal transmission of phenotypes $A$ and $B$, respectively.

@@

%\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Results
%\newpage
\section{Results}

@@

It is interesting to examine the general effect of interaction-transmission association $\alpha$ on the evolution of cooperation.
Define the interaction-transmission association thresholds, $a_1$ and $a_2$, as 
\begin{equation} \label{eq:boundries_assortative_meeting_general_case}
  a_1 = \frac{c\cdot v(1-T_A) -(T_A-T_B)(1+b-c)}{b\cdot v \cdot T_B}, \quad
  a_2 = \frac{c\cdot v(1-T_B)-(T_A-T_B)}{b\cdot v\cdot T_A} \;.
\end{equation}
\\

@@

\begin{figure}[p]
  \centering       
    \includegraphics[width=0.9\textwidth]{../PRSB_figures/fig_3.pdf}
    \caption{\textbf{Evolution of cooperation under vertical, oblique, and horizontal cultural transmission.} 
    The figure shows parameter ranges for global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable polymorphism (blue), and stable polymorphism of cooperation and defection (yellow).
	In all cases the vertical transmission rate $v$ is on the x-axis.
	(\textbf{a-b})~Cost of cooperation $c$ is on the y-axis and the cost thresholds $\gamma_1$ and $\gamma_2$ (Eqs.~\ref{eq:cost_thresholds_v_threshold}) are represented by the solid and dashed lines, respectively. 
    (\textbf{c-d})~Interaction-transmission association $\alpha$ is on the y-axis and the interaction-transmission association thresholds $a_1$ and $a_2$ (Eqs.~\ref{eq:boundries_assortative_meeting_general_case}) are represented by the solid and dashed lines, respectively. 
    Horizontal transmission is biased in favor of cooperation, $T_A>T_B$, in (\textbf{a}) and (\textbf{c}), or defection, $T_A<T_B$, in (\textbf{b}) and (\textbf{d}).    
    Here, $T_A = 0.5$, and
    (\textbf{a}) $b=1.2$, , $T_B = 0.4$, $\alpha = 0.4$;
    (\textbf{b}) $b=2$, $T_B = 0.7$, $\alpha = 0.7$;
    (\textbf{c}) $b=1.2$, $T_B = 0.4$, $c=0.5$;
    (\textbf{d}) $b=2$, $T_B = 0.7$, $c=0.5$.
    }
    \label{fig:equilibria}
\end{figure}

@@

\begin{figure}[p]
  \centering       
	\includegraphics[width=0.9\textwidth]{../PRSB_figures/fig_4.pdf}
    \caption{\textbf{Evolution of cooperation under vertical and horizontal cultural transmission (v=1).} 
    The figure shows parameter ranges for global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable polymorphism (blue), and stable polymorphism of cooperation and defection (yellow).
    (\textbf{a-c})~The horizontal transmission bias ($T_A-T_B$) is on the x-axis.
    In panels \textbf{(a)} and \textbf{(b)}, the cost of cooperation $c$ is on the y-axis and the cost thresholds $\gamma_1$ and $\gamma_2$ (\autoref{eq:cost_thresholds_v_threshold}) are the solid and dashed lines, respectively. 
	In panel \textbf{(c)}, interaction-transmission association $\alpha$ is on the y-axis and the interaction-transmission association thresholds $a_1$ and $a_2$ (Eqs.~\ref{eq:boundries_assortative_meeting_general_case}) are the solid and dashed lines, respectively.
    Here, $b=1.3$, $T_A=0.4$, $v=1$, (a)~$\alpha = 0$, (b)~$\alpha = 0.7$, (c)~$c = 0.35$.    
    \textbf{(d)}~Change in frequency of cooperation among juveniles ($\hat{p}'-\hat{p}$) as a function of the frequency ($\hat{p}$), see \autoref{eq:nextgen_juveniles}.
    The orange curve shows convergence to a stable polymorphism ($T_A = 0.4$, $T_B = 0.9$, $b = 12$, $c=0.35$, $v=1$, and $\alpha = 0.45$). % todo ref eq 12
    The blue curve shows fixation of either cooperation or defection, depending on the initial frequency ($T_A = 0.5$, $T_B = 0.1$, $b = 1.3$, $c=0.904$, $v=1$, and $\alpha = 0.4$).
    Black circles show the three equilibria.
  	}
    \label{fig:equilibria_v1}
\end{figure}
\newpage

@@

\begin{figure}[h]
  \centering       
    \includegraphics[scale=0.5]{../PRSB_figures/fig_5.pdf}
    \caption{\textbf{Unstable and stable equilibrium dynamics.} 
    The figure shows the frequency of cooperation over time (generations) starting from different initial frequencies.
    \textbf{(a)} The system has an unstable equilibrium and therefore all the curves are moving away from it.
    \textbf{(b)} The system has a stable equilibrium and therefore all the curves moves towards it. 
    Here, $T_A = 0.5$, $c = 0.5$,  $v=0.59$ and \textbf{(a)} $T_B = 0.4$, $\alpha=0.1$ and $b=1.2$. \textbf{(b)} $T_B=0.7$, $\alpha=0.6$ and $b=2.3$.
    }
    \label{fig:time_series}
\end{figure}

@@

%%% Figure: Invasion
\begin{figure}[h]
  \centering
  \includegraphics[width=1\textwidth]{../PRSB_figures/fig_6.pdf}
  \caption{
  \textbf{Reduction principle for interaction-transmission association.} 
  Consecutive fixation of modifier alleles that reduce interaction-transmission association $\alpha$ in numerical simulations of evolution with two modifier alleles (Eq.~D1).
  When an invading modifier allele is established in the population (frequency > 99.95\%), a new modifier allele that reduces interaction-transmission association by 5\% is introduced (at initial frequency 0.5\%).
  \textbf{(a)}~The frequency of the cooperative phenotype $A$ over time.
  \textbf{(b)}~The frequency of the invading modifier allele $m$ over time. 
  \textbf{(c)}~The population mean fitness ($\bar{w}$) over time.
  Here, $v=1$, $c = 0.05$, $b=1.3$, $T_A=0.4<T_B=0.7$, initial interaction-transmission association $\alpha_1=0.7$, lower interaction-transmission association threshold $a_2=0.605$.  
  }
  \label{fig:invasion}
\end{figure}
%\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Population structure}
Here, we are going to run stochastic simulations to learn more about evolution of Cooperation when the population is structured. 
The model is described in \autoref{sec:spatialModel}.
All the simulations in this section were made by Ohad Lewin-Epstein from Tel Aviv University.
The outcomes of stochastic simulations with such a structured population are shown in \autoref{fig:spatial}, which demonstrates that the highest cost of cooperation $c$ that permits the evolution of cooperation agrees with the conditions derived above for our model without population structure or stochasticity.
An example of stochastic stable polymorphism is shown in \autoref{fig:spatial}c.
Changing the simulation so that selection is local (i.e. sites can only be settled by offspring of neighboring parents) had only a minor effect on the agreement with the derived conditions (\autoref{fig:spatial_local_selection}).

@@

%\newpage
